---
title: "02-DESeq2 Analysis"
author: "Hannia Larino"
date: "2025-05-27"
output: html_document
---
This script has all of my code for the analysis using DESeq, BLAST, and Uniprot to ID mapping. The table from the data frame called "functional.merge" at chunk 226 has all of the data that came from DESeq, BLAST, and Uniprot to ID mapping. 

## Part1. **RUNNING DESeq2**

*First, pre-process data!* Reading in countmatrix

```{r}
countmatrix <- read.table("~/SeaCucumber/output/kallisto_01.isoform.counts.matrix", header = TRUE, row.names = 1)
head(countmatrix)
```

Filtering according to smallest treatment size of 3.

```{r}
# Filter: keep rows with counts >5 in at least 3 samples
keep <- rowSums(countmatrix > 5) >= 3
filtered.counts <- countmatrix[keep, ]

# Summary
cat("Before filtering:", nrow(countmatrix), "isoforms\n")
cat("After filtering:", nrow(filtered.counts), "isoforms\n")
```

Rounding up data to nearest whole number bc DESeq2 does not take decimals

```{r}
filtered.counts <- round(filtered.counts, 0)
str(filtered.counts)
```

## below table shows counts but rounded to nearest integer

```{r}
filtered.counts
```

Loading libraries to be used for DESeq2

```{r}
library(DESeq2)
library(tidyverse)
library(pheatmap)
library(RColorBrewer)
library(data.table)
```

Creating vector that renames file ID to meaningful ID

```{r}
# Vector of new sample names in the correct order
new.names <- c("Control 1", "Control 2", "Control 3", "26.a1", "26.a2", "26.a3", "26.b1", "26.b2", "26.b3", "30.a1","30.a2","30.a3")
# Apply new names to the columns
colnames(filtered.counts) <- new.names
colnames(filtered.counts)
head(filtered.counts)
```

Telling R we have 3 conditions and the control.

```{r}
condition <- factor(rep(c("Control", "26.a", "26.b", "30.a"), each = 3),
                    levels = c("Control", "26.a", "26.b", "30.a"))
condition
```


```{r}
deseq2.colData <- data.frame(condition = condition, 
                             type=factor(rep("paired", 12)))

rownames(deseq2.colData) <- colnames(filtered.counts)

deseq2.data <- DESeqDataSetFromMatrix(countData = filtered.counts,
                                     colData = deseq2.colData, 
                                     design = ~ condition)
```

Finally getting the DGE analysis from deseq2.res

```{r}
deseq2.data <- DESeq(deseq2.data)
deseq2.res <- results(deseq2.data)
deseq2.res <- deseq2.res[order(rownames(deseq2.data)), ]
```

explore the results from below chunk

```{r}
# Count number of hits with adjusted p-value less then 0.05
dim(deseq2.res[!is.na(deseq2.res$padj) & deseq2.res$padj <= 0.05, ])

head(deseq2.res[order(deseq2.res$padj), ])  # sorted by adjusted p-value
```



VISUALIZATION BELOW

tmp=the main plot

```{r}
tmp <- deseq2.res
# The main plot
plot(tmp$baseMean, tmp$log2FoldChange, pch=20, cex=0.45, ylim=c(-3, 3), log="x", col="darkgray",
     main="DEG of 2 temperatures  (pval <= 0.05)",
     xlab="mean of normalized counts",
     ylab="Log2 Fold Change")

# Getting the significant points and plotting them again so they're a different color
tmp.sig <- deseq2.res[!is.na(deseq2.res$padj) & deseq2.res$padj <= 0.05, ]
points(tmp.sig$baseMean, tmp.sig$log2FoldChange, pch=20, cex=0.45, col="red")
# 2 FC lines
abline(h=c(-1,1), col="blue")
```

## Part 2. Merge DESeq and BLAST data. 

## CREATING TABLE TO MERGE W BLAST

```{r}
write.table(tmp.sig, "~/SeaCucumber/output/DEGlist.tab", row.names = T)
```

## read in table

```{r}
DEGtable <- read.table("~/SeaCucumber/output/DEGlist.tab",
                  header = TRUE,
                  sep = " ",
                  quote = "\"",
                  stringsAsFactors = FALSE,
                  check.names = FALSE)

```

```{r}
DEGtable #NOTE THIS TABLE ONLY SHOWS THE DEG WITH PVAL < 0.05
```

## adding transcript id as a column

```{r}
# Adding transcript_id column to corresponding row 
DEGtable$transcriptID <- rownames(DEGtable)

#moving this column to the beginning of DEGtable
DEGtable <- DEGtable[, c("transcriptID", setdiff(names(DEGtable), "transcriptID"))]


```

```{r}
DEGtable
```

## read in blast

```{r}
blastresults <- read.table("~/SeaCucumber/output/Seacuc_blastx_sprotresults.tab", 
                    header = TRUE, sep = "\t", stringsAsFactors = FALSE)

blast.top <- blastresults[!duplicated(blastresults$transcript_id), ]

# Assign column names to BLAST 
colnames(blastresults) <- c("transcriptID", "subject_id", "percent_identity", "alignment_length",
                     "mismatches", "gap_opens", "q_start", "q_end", "s_start", "s_end", "evalue", "bit_score")

#removing unnecessary stuff from subject_id: not working
#blastresults$gene_symbol <- sub(".*\\|(.*)_.*", "\\1", blastresults$subject_id)

```

```{r}
head(blastresults)
colnames(blastresults$gene_symbol)
```

## MERGE BLAST and DEG LIST

```{r}
deg.annotated <- merge(DEGtable, blastresults, by = "transcriptID", all.x = TRUE)
write.csv(deg.annotated, "~/SeaCucumber/output/deg_annotated.csv", row.names = FALSE)
#made the csv for the presentation
```

## Part 3: Functional analysis

## creating text file of subject_id to paste into <https://www.uniprot.org/id-mapping>

```{r}
deg.annotated$accession <- sub("sp\\|(.*?)\\|.*", "\\1", deg.annotated$subject_id)
write.table(deg.annotated$accession, file = "~/SeaCucumber/output/uniprot_ids.txt", quote = FALSE, row.names = FALSE, col.names = FALSE)
```

Used <https://www.uniprot.org/id-mapping> to paste the .txt file with the subject_ids.

## opening file

```{r}
uniprot.annot <- read.delim("~/SeaCucumber/output/uniprot_idmapping.tsv.gz", header = TRUE, sep = "\t", stringsAsFactors = FALSE)
```

```{r}
colnames(uniprot.annot)
```

```{r}
deg.annotated$accession <- sub("sp\\|(.*?)\\|.*", "\\1", deg.annotated$subject_id)
deg.annotated
```

## merging uniprot data with blast/seq table

```{r}
functional.merge <- merge(deg.annotated, uniprot.annot, by.x = "accession", by.y = "From", all.x = TRUE)
write.csv(functional.merge, "~/SeaCucumber/output/functional_merge.csv", row.names = FALSE) #for slides
```
#### MAIN DATA TABLE BELOW: functional.merge
## This data table is a compilation of DEseq2 resutls (p-val \</ 0.05), BLAST to Swiss-Prot results, and Uniprot ID mapping results.

```{r}
functional.merge
```

## enrichment analysis (final analysis) using gprofiler2 package to use gost()

```{r}
deg_ids <- unique(deg.annotated$accession) #getting accession ID (used for uniprot id mapping)
```

```{r}
install.packages("gprofiler2")
```

```{r}
library(gprofiler2)
# Use cleaned UniProt accessions of DEGs
gost.res <- gost(query = deg_ids, organism = "hsapiens") #using hsapiens as organism lol

```

```{r}
# View results
head(gost.res)

```

```{r}
# Select top 10 significant terms
top_terms <- head(gost.res$result[order(gost.res$result$p_value), ], 10)

plot<- ggplot(top_terms, aes(x = reorder(term_name, -log10(p_value)), 
                      y = -log10(p_value), 
                      fill = source)) +
  geom_col() +
  coord_flip() +
  labs(title = "Top Enriched Terms", x = "GO/Pathway Term", y = "-log10(p-value)") +
  theme_minimal()

ggsave("~/Hannia-jelly/assignments/images/top_enriched_terms.png", plot = plot, width = 8, height = 5, dpi = 300)
```
## end of project, see compendium for analysis of data. 

