---
title: "03-Blast md"
output: html_document
date: "2025-04-17"
output: 
  html_document:
    theme: readable
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
---

**HOW TO COMPLETE BLAST**

1. Downloading latest NCBI software .gz file from https://ftp.ncbi.nlm.nih.gov/blast/executables/blast+/LATEST/

```{r, engine='bash'}

cd /home/jovyan/applications #setting directory 
curl -O https://ftp.ncbi.nlm.nih.gov/blast/executables/blast+/LATEST/ncbi-blast-2.16.0+-x64-linux.tar.gz #uploading NCBI software to directory 
tar -xf ncbi-blast-2.16.0+-x64-linux.tar.gz #extract files from .tar archive 

```
**Check to see if the software downloaded in the "applications" folder.**
```{r, engine='bash'}
ls /home/jovyan/applications/ncbi-blast-2.16.0+-x64-linux.tar.gz
```
**MAKING THE DATABASE**
2. Make a blast database. Download latest Swiss-Prot protein sequences.
```{r, engine='bash'}
curl -O https://ftp.uniprot.org/pub/databases/uniprot/current_release/knowledgebase/complete/uniprot_sprot.fasta.gz
#curl -O saves output to a file. This file is an unzipped folder. 

mv uniprot_sprot.fasta.gz uniprot_sprot_r2023_01.fasta.gz
#mv used to move or rename files and directories 
#accidentally renamed 2023 instead of 2025 

gunzip -k uniprot_sprot_r2025_01.fasta.gz
#gunzip -k decompresses file but preserves original .gz file. 
ls 
#lists files and directories in current directory, when run ls w/out argument, it displays names of files and folders in current directory 
```
3. Uniprot files downloaded in blastdb folder. Use 'head' to see file. 
```{r, engine='bash'}
#checking to see first lines of file 
cd /home/jovyan/blastdb
head /home/jovyan/blastdb/uniprot* 
```
4. This actually creates the BLAST database in the blastdb folder. 
```{r, engine='bash'}
/home/jovyan/applications/ncbi-blast-2.16.0+/bin/makeblastdb \
-in ../../blastdb/uniprot_sprot_r2023_01.fasta \
-dbtype prot \
-out /home/jovyan/blastdb/uniprot_sprot_r2023_01

```

**GETTING QUERY SEQUENCE**
5. Download the FASTA file "Ab_4denovo_CLC6_a.fa" 
```{r, engine='bash'}
cd /home/jovyan/applications/data
curl https://eagle.fish.washington.edu/cnidarian/Ab_4denovo_CLC6_a.fa \
-k \
> /home/jovyan/applications/data/Ab_4denovo_CLC6_a.fa 
```
**View head of Ab_4denovo**
```{r, engine='bash'}
head /home/jovyan/applications/data/Ab_4denovo_CLC6_a.fa

```
**insert note** 
```{r, engine='bash'}
echo "How many sequences are there?"
grep -c ">" /home/jovyan/applications/data/Ab_4denovo_CLC6_a.fa

#grep -c Counts the number of lines that start with > — which correspond to sequence headers in a FASTA file. This gives you the number of sequences in the file.
```
**RUN BLAST** 

```{bash}
pwd
cd /home/jovyan/applications
pwd 
```

Getting an error here 
Warning: [blastx] Examining 5 or more matches is recommended
BLAST Database error: Database memory map file error
```{r, engine='bash'}
cd /home/jovyan/applications
/home/jovyan/applications/ncbi-blast-2.16.0+/bin/blastx \
-query /home/jovyan/applications/data/Ab_4denovo_CLC6_a.fa \
-db ../../blastdb/uniprot_sprot_r2023_01 \
-out output/Ab_4-uniprot_blastx.tab \
-evalue 1E-20 \
-num_threads 1 \
-max_target_seqs 1 \
-outfmt 6

```
```{bash}

head -2 output/Ab_4-uniprot_blastx.tab
#-2 part says you only want to see first two lines of file 
wc -l output/Ab_4-uniprot_blastx.tab
#wc = word count, -1 = counts number of files 

```

